Stability in human interaction networks: primitive typology of vertex, prominence of measures and activity statistics

### 2015-01-21 paper.tex
terminado de incorporar os apontamentos do chu

\section{Data description}
    \subsection{Email lists and messages}
Email list messages were obtained from
the GMANE email archive~\cite{GMANE}, which consists of more than 20,000 email lists and more than 130,000,000 messages~\cite{GMANEwikipedia}. These lists cover a variety of topics, mostly technology-related. It can be described as a corpus with metadata of its messages, such as time, place, sender name, sender email address.
GMANE usage in scientific research is reported in studies of isolated lists and of lexical innovations~\cite{GMANE2,bird}. Appendix~\ref{scripts} is dedicated scripts for gathering and processing GMANE email messages.

    \subsection{Chosen dataset}
The four lists below were selected for the resulting diversity, easing initial observance of natural and general properties.
\begin{itemize}
    \item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}. Dominated by participants with hybrid artistic and technological interests. Participants are from different countries, and English is the language used the most. Abbreviated as LAU from now on.
    \item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}. Participants are from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
    \item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}. Dominated by specialized computer programmers. Participants are from different countries, and English is the language used the most. Abbreviated as CPP from now on.
    \item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}. Dominated by Brazilian activists and digital culture interests. Participants are mostly Brazilians, and Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

@@

\begin{table}
  \centering
  \caption{Columns $date_1$ and $date_M$ have first and last message dates from the 20,000 messages considered in each email list.
$N$ is the number of participants (number of different email addresses).
$\Gamma$ is the number of threads (count of messages without antecedent).
$\overline{M}$ is messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
MET notably has the fewer participants and the larger number of threads.
This relation holds for each pair of the lists considered: as the number of participants increases,
 the number of threads decreases.}
  \label{geralListas}
  \begin{tabular}{|l|c|c|c|c|c|}\hline
list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
LAU  & Jun/29/2003 & Jul/23/2005 & 1183 & 3373 & 5 \\
LAD  & Jun/30/2003 & Oct/07/2009 & 1268 & 3113 & 4 \\
MET  & Ago/01/2005 & Mar/07/2008 & 492  & 4607 & 23 \\
CPP  & Mar/13/2002 & Aug/25/2009 & 1052 & 4506 & 7 \\ \hline
  \end{tabular}
\end{table}

@@

    \subsection{Interaction network}\label{intNet}
Regarding literature~\cite{bird,newmanCommunityDirected,newmanCommunity2013}, interaction networks can be modeled both weighted or unweighted, both directed or undirected. Networks in this article are directed and weighted, considered as more informative among possibilities (directed unweighted, undirected weighted, and undirected unweighted).
More precisely, the networks reported are erected as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network, as B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to edge weight. Self-loops were regarded as non-informative and discarded. This networks are described as exhibiting free-scale and small world properties, as expected for a social network~\cite{bird}.

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/criaRede_}
    \caption{Formation of interaction network from email messages. Each vertex represents a participant. A reply message from participant B to a message from participant A is considered as evidence that B received information from A. Multiple messages add ``weight'' to directed edge. Further details are in Section~\ref{intNet}.}
    \label{formationNetwork}
\end{figure}

Previous messages on the thread create directed edges from their author to the observed message's author. Edges can be created from all antecedent messages on the message-response thread. In this work, only the immediate predecessor are linked to new message's author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a connection between $x\rightarrow z$. Potential interpretations for this weaker connection are usually common sense, such as: double length, half weight or with one more ``obstacle''. This suggests the adoption of other centrality measures that account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

@@

\begin{equation}
    P(k)=\binom{2(N-1)}{k}p^k(1-p)^{2(N-1)-k}
\end{equation}

The lower degree fat tail constitute the border
vertexes or peripheral sector. 
The higher degree fat tail is the hub sector.

The arguments behind this classification are: 1) vertexes so connected that they are virtually inexistent in networks whose edges are incident regardless of vertex properties, are correctly associated to the hubs sector. Vertexes with very few connections, which are way more abundant than expected in networks whose edges are incident regardless of vertex properties, are correctly associated to periphery. Degree values near average, predicted as the most abundant if connections are created without criteria, and less frequent in free-scale phenomena, are correctly associated to intermediary vertexes.

To assure statistical validity, bins can be chosen to contain at least $\eta$ vertexes. Thus, each bin, starting at degree $k_i$, spans $\Delta_i=[k_{i},k_{j}]$ degree values, where $j$ is the smallest integer in which degrees $k_i$-$k_{j}$ contain at least $\eta$ values. This changes equation~\ref{criterio} to:

@@

If instead strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}$ the average weight of an edge and $s_i$ the vertex strength. For in and out degrees and strengths, comparisons should be made with $\kappa_i=2k_i^{in}$, $\kappa_i=2k_i^{out}$, $\kappa_i=2\frac{s_i^{in}}{\overline{w}}$ and $\kappa_i=2\frac{s_i^{out}}{\overline{w}}$. Results of these criteria for network segmentation are discussed in subsection~\ref{subsec:pih}.

As a further refinement of the network segmentation, compound criteria is used for classification of vertex, considering all measures: total, in and out degree and strength. After a careful inspection of possible combinations, these were abbreviated to six:

\begin{itemize}
    \item Exclusivist criteria: vertex are only classified if the class is the same with respect to all measures. In this case, total vertex classified (usually) does not reach 100\%, which is depicted by a black line in Appendix~\ref{figures}.
    \item Inclusivist criteria: vertex has class given by any of the measures. In this case, a vertex can have more than one class and the fraction of class attribution beyond total number of vertexes is also depicted by a black line in Appendix~\ref{figures}.
    \item Exclusivist cascade: hubs are only hubs if classified as hub with respect to all measures. Intermediary are the vertexes classified as intermediary or hub with respect to all measures. Vertexes left are regarded as peripheral vertex.
    \item Inclusivist cascade: hubs are vertexes classified as hubs by any of the measures. From the vertexes left, if any is classified as intermediary by any measure, than it is intermediary. The rest of the vertexes are peripheral.
    \item Exclusivist externals: hubs have unanimous classification with respect to all measures. Of vertexes left, peripheral vertexes are the ones classified as hub or peripheral by simple criterion. The rest represent intermediary sector.
    \item Inclusivist externals: hubs are vertexes classified as hubs with respect to any measure. From vertexes left, if a vertex is classified as peripheral with respect to any measure, than it is peripheral. The rest is regarded intermediary sector.
\end{itemize}

These compound criteria, and reduction of possibilities to them, can be formalized in strict mathematical terms. This was considered out of the scope of the present article.

Results from applying this classification method is further reported in Section~\ref{subsec:pih}.

        \subsubsection{Topological measures}\label{measures}
This article
restricts topological analysis to a small selection of the most basic
measures of each vertex:

\begin{itemize}
    \item Degree $d_i$: number edges linked to node $i$.
    \item In-degree $d_i^{in}$: number of edges ending at node $i$.
    \item Out-degree $d_i^{out}$: number of edges departing from node $i$.
    \item Strength $s$: sum of weights of all edges linked to node $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at node $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from node $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked. This is the standard clustering coefficient for undirected graphs. Usage of other clustering coefficients (e.g. for directed graphs) was considering out of scope, as such clustering coefficients are usually employed in special contexts. 
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain the node $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

In order to capture asymmetries in the activity of participants, the following metrics were introduced (see subsection~\ref{prevalence}):

\begin{itemize}
    \item asymmetry of vertex $i$: $asy_i=\frac{d_i^{in}-d_i^{out}}{d_i}$.
    \item mean of edge asymmetry of vertex $i$: $\mu_i^{asy}=\frac{\sum_{j\in J_i} e_{ji}-e_{ij}}{|J_i|}$. Where $e_{xy}$ is 1 if there is and edge from $x$ to $y$, $0$ otherwise. $|J_i|$ is the number of neighbors of vertex $i$.
    \item standard deviation of edge asymmetry of vertex $i$: $\sigma_i^{asy}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{asy} -(e_{ji}-e_{ij}) ]^2  }{|J_i|}  }$
    \item disequilibrium of vertex $i$: $dis_i=\frac{s_i^{in}-s_i^{out}}{s_i}$.
    \item mean of edge disequilibrium of vertex $i$: $\mu_i^{dis}=\frac{\sum_{j \in J_i}\frac{w_{ji}-w_{ij}}{s_i}}{|J_i|}$, where $w_{xy}$ is weight of edge $x\rightarrow y$ and zero if there is no such edge.
    \item standard deviation of edge disequilibrium of vertex $i$: $\sigma_i^{dis}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{dis}-\frac{(w_{ji}-w_{ij})}{s_i}]^2}{|J_i|}}$
\end{itemize}

   \subsection{Evolutive observations}
Evolution of network is observed within a fixed number of messages (window size: $ws$) that shifts in the message timeline.
 All $ws=$ 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000 were used. Within a same $ws$, the number vertexes and edges vary in time, as do other network characteristics. Further work should deepen inspection of measure interdependence, this article holds to measures in Section~\ref{measures}.

        \subsubsection*{Visualization of network evolution}
In refining hypotheses, visualization of the network was crucial. Animations, image galleries and online gadgets were made~\cite{animacoes,galGMANE,appGMANE}. Mapping of various topological measures to glyphs and layouts are being further explored as a parallel research. Furthermore, stable aspects of measures prominence along time are captured through mean and standard deviation (see Section~\ref{measures} and Appendix~\ref{sec:pcat}). Constant sector sizes along time are observed in a timeline fashion in Appendix~\ref{figures}.

@@

\section{Results and discussion}
    \subsection{Constancy and discrepancy of activity along time}\label{constDisc}
        \subsubsection{Seconds and minutes}
The incidence of messages at each second in a minute and at each minute in an hour is compatible with uniform distribution tests.
If compared to simulations using an uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function.}, messages were slightly more evenly distributed in all lists: for both seconds and minutes  $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$. Simulations reach these values, but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.

        \subsubsection{Hours of the day}
Table~\ref{dia} shows how the four lists distribute activity along the day.
Afternoon was the most active 6h period of the day. Second 12h more active than first 12h. Even so, activity peak occurs around midday, with a skew towards earlier hours.

        \subsubsection{Days of the week}
 Weekdays also exhibit an interesting pattern, to which is dedicated Table~\ref{semana}: a decrease of activity on weekends of at least one third, reaching two thirds in extreme cases.
        \subsubsection{Days along the month}
 Table~\ref{mes} shows activity along the month, in which no prevalent pattern was observed.
 Variation of activity in the days along the month is less prominent, one cannot point much more than a - probably not statistically relevant - tendency of first and second weeks to be more active. The most important trait might be their homogeneity with respect to activity. Last days of the month (29, 30 and 31) are not present in every month, and observed activity is proportional to incidence rates.

@@

As specified in Section~\ref{sectioning}, in order to classify nodes as hubs, intermediary or peripheral, the incident connectivity distribution $\widetilde{P}(k)$ is compared
to a connectivity distribution of an ideal Erd\"os R\`enyi network $P(k)$ with the same number of vertexes and edges. In the networks inspected for this article (see Section~\ref{intNet}), if degree distribution is used for classification, hubs sector size reaches peaks with $10\%$ of all vertexes. 
 If strength is used for comparison ($\widetilde{P}(k_i=\frac{s_i}{\overline{w}})$) is compared
to $P(k_i)$), hubs account for approximately $5\%$ of all vertex, i.e. strength classification yields half the number of hubs as plain degree. 
This results hold for in and out degrees and strengths,

Classification criteria exposed in Section~\ref{sectioning} was used efficiently with windows of at least 200 messages. Specially with 1000 or more messages, criteria yields stable fractions of $\approx 5\%$ of hubs, $\approx [15-20]\%$ of intermediary and $\approx [75-80]\%$ peripheral vertexes, which match literature expectations~\cite{secFree}. A compound criteria, also described in Section~\ref{sectioning}, can be used as a classification refinement. This is specially useful in dealing with fewer messages, in which case the structure degenerates with respect to some of the degree and strength measures, but not all.
A minimum window size for observation of more general properties can be inferred by monitoring the giant component and the degeneration of hub, intermediary and peripheral sections. This degeneration is critical in the span of 50-100 messages for window sizes. With compound criteria, such as exclusive cascade of Figure~\ref{fig:cpp250_}, the networks seem to hold basic structure even with as few as 20-50 messages. This indicates that concentration of activity and low-activity participants occurs even with very few messages.

@@

The principal component (PCA~\cite{pca}) exhibit ponderation of centrality measures: degrees, strengths and betweenness centrality. Clustering coefficient is presented in almost perfect orthogonality. Dispersion is more prevalent in symmetry related measures than clustering coefficient. This holds for all network snapshots observed, even with as few messages as to degenerate structure. Symmetric and asymmetric edges have been reported as bonded to different roles played by participants and relations~\cite{newmanEvolving}. Principal components formation from original measures can be observed in Tables~\ref{compPCA0},~\ref{compPCA} and~\ref{compPCA2}. Individual vertexes relation with top two principal components can be appreciated in Figures~\ref{PCA} and~\ref{PCA2}.
This peculiar first component that consists of the averaged sum of degree, strength and betweenness measures was verified to be incident in virtually all networks with 500 or more messages and most smaller networks (degeneration of basic structure is critical with $ws \approx 50-100$ messages). This composition of principal component suggests that all six degree and strength measures are equally important for system characterization, although it is known that they do not relate to the same participation characteristics.

As expected, degree and strength are highly correlated, with Spearman correlation coefficient $\in [0.95,1]$ and Pearson coefficient $\in [0.85,1)$ for larger window sizes ($ws>1000$). Also trivial: high degree is associated with low clustering coefficient, as can be observed in Figure~\ref{clust}.

When symmetry of node connectivity is considered, the first component remains mostly the same, but clustering coefficient is only relevant to third and fourth components. A snapshot of vertexes with respect to these first two principal components are in Figure~\ref{PCA2}. This asymmetry and disequilibrium measures revealed as more proper measures to characterization of hubs and intermediaries, as seen in greater spreading of second PCA plot. Both symmetry of node overall activity and of individual relations play important role in second component, as can be observed in Table~\ref{compPCA2}.

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr3PCA}
    \caption{PCA of in and out degree and strength, betweenness centrality and clustering coefficient, as specified in Section~\ref{measures}. Table~\ref{compPCA} has the composition of principle components from the original measures. First principle component is a pondered sum of centrality measures: degrees, strengths and betweenness centrality. Second component is mostly clustering coefficient in this figure, but asymmetries holds second component if also considered. Similarity to plot in Figure~\ref{clust} was verified with all window sizes considered ($ws\;\in\;[100,10000]$), which exposes a common relation is held by degree, strength and betweenness measures to clustering coefficient.}
    \label{PCA}
\end{figure}

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr1PCA}
    \caption{Degree and strength, clustering coefficient, betweenness centrality and symmetry related measures are used for this scatter plot of principal components. Compositions of first three components are in table~\ref{compPCA2} and measure details in subsection~\ref{measures}. Most importantly, clustering coefficient is only relevant for third component, being second component representative of symmetry measurements of vertex interactions.}
    \label{PCA2}
\end{figure}

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr11CC}
    \caption{Clustering versus degree of vertex with a window size of $ws = 1000$ email messages (LAU list). General layout is in accordance with literature: connected vertexes have low clusterization while higher clusterization is gradually more incident as number of connections is lowered.}
    \label{clust}
\end{figure}

@@

Regarding topological aspects of interaction networks, further work should inspect other measures (e.g. closeness centrality, accessibility), and statistics in each of the three connective sectors: hubs, intermediary and peripheral.

>>>>

\section{Data description}
    \subsection{Email lists and messages}
Email list messages were obtained from
the GMANE email archive~\cite{GMANE}, which consists of more than 20,000 email lists and more than 130,000,000 messages~\cite{GMANEwikipedia}. These lists cover a variety of topics, mostly technology-related. The archive can be described as a corpus with metadata of its messages, including sent time, place, sender name, sender email address.
GMANE usage in scientific research is reported in studies of isolated lists and of lexical innovations~\cite{GMANE2,bird}. The scripts for gathering and processing GMANE email messages are given in Appendix~\ref{scripts}.

    \subsection{Chosen dataset}
Four email lists were selected for their diversity, easing initial observance of natural and general properties.
\begin{itemize}
    \item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}. Dominated by participants with hybrid artistic and technological interests. Participants are from different countries, and English is the language used the most. Abbreviated as LAU from now on.
    \item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}. Participants are from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
    \item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}. Dominated by specialized computer programmers. Participants are from different countries, and English is the language used the most. Abbreviated as CPP from now on.
    \item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}. Dominated by Brazilian activists and digital culture interests. Participants are mostly Brazilians, and Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

@@

\begin{table}
  \centering
  \caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
$N$ is the number of participants (number of different email addresses).
$\Gamma$ is the number of threads (count of messages without antecedent).
$\overline{M}$ is the number of messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
MET notably has the fewer participants and the larger number of threads.
This relation holds for each of the lists considered: as the number of participants increases,
 the number of threads decreases.}
  \label{geralListas}
  \begin{tabular}{|l|c|c|c|c|c|}\hline
list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
LAU  & Jun/29/2003 & Jul/23/2005 & 1183 & 3373 & 5 \\
LAD  & Jun/30/2003 & Oct/07/2009 & 1268 & 3113 & 4 \\
MET  & Ago/01/2005 & Mar/07/2008 & 492  & 4607 & 23 \\
CPP  & Mar/13/2002 & Aug/25/2009 & 1052 & 4506 & 7 \\ \hline
  \end{tabular}
\end{table}

@@

    \subsection{Interaction network}\label{intNet}
Regarding literature~\cite{bird,newmanCommunityDirected,newmanCommunity2013}, interaction networks can be modeled both weighted or unweighted, both directed or undirected. Networks in this article are directed and weighted, considered as more informative among possibilities (directed unweighted, undirected weighted, and undirected unweighted).
The networks were considered as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network, as B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network as described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to edge weight. Self-loops were regarded as non-informative and discarded. This networks are described as exhibiting free-scale and small world properties, as expected for a social network~\cite{bird}.

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/criaRede_}
    \caption{Formation of interaction network from email messages. Each vertex represents a participant. A reply message from participant B to a message from participant A is regarded as evidence that B received information from A. Multiple messages add ``weight'' to directed edge. Further details are in Section~\ref{intNet}.}
    \label{formationNetwork}
\end{figure}

Previous messages on the thread create directed edges from their author to the observed message's author. Edges can be created from all antecedent messages on the message-response thread. In this work, only the immediate predecessor is linked to new message's author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a connection between $x\rightarrow z$. Potential interpretations for this weaker connection are usually common sense, such as: double length, half weight or with one more ``obstacles''. This suggests the adoption of centrality measures that account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

@@

\begin{equation}
    P(k)=\binom{2(N-1)}{k}p_e^k(1-p_e)^{2(N-1)-k}
\end{equation}

The lower degree fat tail is constituted by the border
vertexes and is called the peripheral sector. 
The higher degree fat tail is the hub sector.

The arguments behind this classification are: 1) vertexes so connected that they are virtually inexistent in networks whose edges are incident regardless of vertex properties, are correctly associated to the hubs sector. Vertexes with very few connections, which are way more abundant than expected in networks whose edges are incident regardless of vertex properties, are correctly assigned to periphery. Degree values near average, predicted as the most abundant if connections are created without criteria, and less frequent in free-scale phenomena, are classified to intermediary vertexes.

To ensure statistical validity, bins can be chosen to contain at least $\eta$ vertexes. Thus, each bin, starting at degree $k_i$, spans $\Delta_i=[k_{i},k_{j}]$ degree values, where $j$ is the smallest integer in which degrees $k_i$-$k_{j}$ contain at least $\eta$ vertexes. This changes equation~\ref{criterio} to:

@@

If instead strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}=\frac{z}{\sum_is_i}$ the average weight of an edge and $s_i$ the strength of vertex $i$. For in and out degrees and strengths, comparisons should be made with $\kappa_i=2k_i^{in}$, $\kappa_i=2k_i^{out}$, $\kappa_i=2\frac{s_i^{in}}{\overline{w}}$ and $\kappa_i=2\frac{s_i^{out}}{\overline{w}}$. Results of these criteria for network segmentation are discussed in Section~\ref{subsec:pih}.

Compound criteria is also used for classification of vertex, considering all measures: total, in and out degree and strength. After a careful inspection of possible combinations, these were reduced to six:

\begin{itemize}
    \item Exclusivist criteria: vertexes are only classified if the class is the same according to all metrics. In this case, vertexes classified does not usually reach 100\%, which is depicted by a black line in Appendix~\ref{figures}.
    \item Inclusivist criteria: vertex has class given by any of the metrics. Therefore, a vertex can belong to more than one class, and class attribution may add more than total number of vertexes, which is indicated by a black line in Appendix~\ref{figures}.
    \item Exclusivist cascade: a hub is only classified as hub if it is a hub with respect to all metrics. Intermediary are the vertexes classified as intermediary or hub with respect to all metrics. Remaining vertexes are regarded as peripheral.
    \item Inclusivist cascade: hubs are vertexes classified as hubs by any of the metrics. The remaining vertexes are classified as intermediary if they belong to this category by any of the metrics. Vertexes only classified as peripheral are the only considered as such.
    \item Exclusivist externals: vertexes are only hubs if they are classified as such according to all the metrics. The remaining vertexes are classified as peripheral if they fall into hub or peripheral categories by any metric. The rest of the vertexes are classified as intermediary.
    \item Inclusivist externals: hubs are vertexes classified as hubs according to any metric. The remaining vertexes will be peripheral if classified as such according to any metric. The rest of the vertexes are regarded regarded as intermediary.
\end{itemize}

These compound criteria, and reduction of possibilities to them, can be formalized in strict mathematical terms. This was considered out of the scope of the present article. Important here is to notice that the compound criteria can be used to examine network sections in the case of a low number of messages, as will be shown later.

Results from applying this classification method is reported in Section~\ref{subsec:pih}.

        \subsubsection{Topological measures}\label{measures}
The topology of the networks was characterized by a small selection of the most standard
measurements of each vertex:

\begin{itemize}
    \item Degree $d_i$: number edges linked to                                 vertex $i$.
    \item In-degree $d_i^{in}$: number of edges ending at                      vertex $i$.
    \item Out-degree $d_i^{out}$: number of edges departing from               vertex $i$.
    \item Strength $s$: sum of weights of all edges linked to                  vertex $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at        vertex $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of vertex $i$ that are linked. This is the standard clustering coefficient for undirected graphs. Usage of other clustering coefficients (e.g. for directed graphs) was considering out of scope, as such clustering coefficients are usually employed in special contexts. 
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain the vertex $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

In order to capture asymmetries in the activity of participants, the following metrics were introduced (see Section~\ref{prevalence}):

\begin{itemize}
    \item Asymmetry of vertex $i$: $asy_i=\frac{d_i^{in}-d_i^{out}}{d_i}$.
    \item Mean of edge asymmetry of vertex $i$: $\mu_i^{asy}=\frac{\sum_{j\in J_i} e_{ji}-e_{ij}}{|J_i|}$. Where $e_{xy}$ is 1 if there is and edge from $x$ to $y$, $0$ otherwise. $|J_i|$ is the number of neighbors of vertex $i$.
    \item Standard deviation of edge asymmetry of vertex $i$: $\sigma_i^{asy}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{asy} -(e_{ji}-e_{ij}) ]^2  }{|J_i|}  }$
    \item Disequilibrium of vertex $i$: $dis_i=\frac{s_i^{in}-s_i^{out}}{s_i}$.
    \item Mean of edge disequilibrium of vertex $i$: $\mu_i^{dis}=\frac{\sum_{j \in J_i}\frac{w_{ji}-w_{ij}}{s_i}}{|J_i|}$, where $w_{xy}$ is weight of edge $x\rightarrow y$ and zero if there is no such edge.
    \item Standard deviation of edge disequilibrium of vertex $i$: $\sigma_i^{dis}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{dis}-\frac{(w_{ji}-w_{ij})}{s_i}]^2}{|J_i|}}$
\end{itemize}

   \subsection{Evolutive observations}
Evolution of network was observed within a fixed number of messages (or window size $ws$) that shifts in the message timeline.
 All $ws=$ 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000 were used. Within a same $ws$, the number vertexes and edges vary in time, as do other network characteristics. Further work should deepen inspection of measure interdependence.

        \subsubsection*{Visualization of network evolution}
In refining hypotheses, visualization of the network was crucial.

The evolution of the networks was visualized with animations, image galleries and online gadgets made for this research~\cite{animacoes,galGMANE,appGMANE}. Mapping of various topological measures to glyphs and layouts are being further explored as a parallel research. Furthermore, stable aspects of measures prominence along time are captured through mean and standard deviation (see Section~\ref{measures} and Appendix~\ref{sec:pcat}). Constant sector sizes along time are observed in a timeline fashion in Appendix~\ref{figures}.

@@

\section{Results and discussion}
    \subsection{Constancy and discrepancy of activity along time}\label{constDisc}
        \subsubsection{Seconds and minutes}
The incidence of messages at each second of a minute and at each minute of an hour is compatible with uniform distribution simulations\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations.}. Messages were slightly more evenly distributed in all lists: for both seconds and minutes  $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$. Simulations reach these values, but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.

        \subsubsection{Hours of the day}
Table~\ref{dia} shows how the four lists distribute activity along the day.
Afternoon was the most active 6h period of the day. Second 12h is $\frac{1}{3}$ more active than first 12h. Even so, activity peak occurs around midday, with a skew toward one hour before noon.

        \subsubsection{Days of the week}
 Weekdays also exhibit an interesting pattern, to which is dedicated Table~\ref{semana}: a decrease of activity on weekends of at least one third, reaching two thirds in extreme cases.

        \subsubsection{Days along the month}
 Table~\ref{mes} shows activity along the month. Variation of activity in the days along the month is less prominent, one cannot point much more than a - probably not statistically relevant - tendency of first and second weeks to be more active. The most important trait seems to be homogeneity. Last days of the month (29, 30 and 31) are not present in every month, and observed activity is proportional to incidence rates.

@@

As specified in Section~\ref{sectioning}, in order to classify nodes as hubs, intermediary or peripheral, the incident connectivity distribution $\widetilde{P}(k)$ is compared
to a connectivity distribution of an ideal Erd\"os R\`enyi network $P(k)$ with the same number of vertexes and edges. In the networks inspected for this article (see Section~\ref{intNet}), if degree distribution is used for classification, hubs can reach $10\%$ of all vertexes. 
 If instead strength is used for comparison ($\widetilde{P}(k_i=\frac{s_i}{\overline{w}})$) is compared
to $P(k_i)$), hubs account for approximately $5\%$ of all vertex, i.e. strength classification yields half the number of hubs as plain degree. This results hold for in and out degrees and strengths.

Classification criteria exposed in Section~\ref{sectioning} reveled remarkable stability over time, provided that a sample of at least 200 messages is taken as the window of analysis. Specially with 1000 or more messages, stable fractions of $\approx 5\%$ of hubs, $\approx [15-20]\%$ of intermediary and $\approx [75-80]\%$ peripheral vertexes were obtained, which match literature expectations~\cite{secFree}.
Compound criterion resulted in a greater stability and a window of only 200 messages was already sufficient for the analysis.
A minimum window size for observation of more general properties can be inferred by monitoring the giant component and the degeneration of hub, intermediary and peripheral sections. This degeneration is critical in the span of 50-100 messages. Upon usage of the compound criteria, such as exclusive cascade of Figure~\ref{fig:cpp250_}, the network seem to hold basic structure even with as few as 20-50 messages. This indicates that concentration of activity and low-activity participants occurs even with very few messages.

@@

The principal component (PCA~\cite{pca}) exhibit ponderation of centrality measures: degrees, strengths and betweenness centrality. Clustering coefficient is presented in almost perfect orthogonality. Dispersion is more prevalent in symmetry related measures than clustering coefficient. Symmetric and asymmetric edges have been reported as bonded to different roles played by participants and relations~\cite{newmanEvolving}. Principal components formation from original measures can be observed in Tables~\ref{compPCA0},~\ref{compPCA} and~\ref{compPCA2}. Individual vertexes relation with top two principal components can be appreciated in Figures~\ref{PCA} and~\ref{PCA2}.
This peculiar first component that consists of the averaged sum of degree, strength and betweenness measures was verified to be incident in virtually all networks with 500 or more messages and most smaller networks (degeneration of basic structure is critical with $ws \approx 50-100$ messages). This composition of the principal component suggests that all six degree and strength measures are equally important for system characterization, although it is known that they do not relate to the same participation characteristics. When symmetry measures are used in PCA, though, the third component has special contributions of out degree and strength.

As expected, degree and strength are highly correlated, with Spearman correlation coefficient $\in [0.95,1]$ and Pearson coefficient $\in [0.85,1)$ for $ws>1000$. Also trivial in the light of complex network literature: high degree is associated with low clustering coefficient, as shown in Figure~\ref{clust}.

When symmetry metrics are included, the first component remains mostly the same, but clustering coefficient is only relevant to third and fourth components. A snapshot of vertexes with respect to these first two principal components is in Figure~\ref{PCA2}. This asymmetry and disequilibrium measures revealed as more proper measures to characterization of hubs and intermediaries, as seen in greater spreading of second plot. These results are summarized in Table~\ref{compPCA2}.

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr3PCA}
    \caption{PCA of in and out degree and strength, betweenness centrality and clustering coefficient, as specified in Section~\ref{measures}. Table~\ref{compPCA} exposes the composition of principle components. Principle component is a pondered sum of centrality measures: degrees, strengths and betweenness centrality. Second component is mostly clustering coefficient. Similarity to plot in Figure~\ref{clust} was verified within all window sizes considered ($ws\;\in\;[100,10000]$), which exposes a common relation is held by degree, strength and betweenness measures to clustering coefficient.}
    \label{PCA}
\end{figure}

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr1PCA}
    \caption{Degree and strength, clustering coefficient, betweenness centrality and symmetry related measures are used for this scatter plot of principal components. Compositions of first three components are in Table~\ref{compPCA2} and measure details in Section~\ref{measures}. Most importantly, clustering coefficient is only relevant for third component, being second component representative of symmetry measurements of vertex interactions.}
    \label{PCA2}
\end{figure}

\begin{figure} 
   \centering
        \includegraphics[width=\columnwidth]{figs/ev0pr11CC}
    \caption{Clustering coefficient versus degree of vertex with a window size of $ws = 1000$ email messages, LAU list. General layout is in accordance with literature: connected vertexes have low clusterization while higher clusterization is gradually more incident as number of connections is lowered.}
    \label{clust}
\end{figure}

@@

Regarding topological aspects of interaction networks, further work should inspect other measures in each of the three connective sectors: hubs, intermediary and peripheral.
